### 2018-06-04 doc/manuscript.Rmd
edits as per Tony's suggestions

Generalized estimating equation (GEE) models [@Zeger1986a] were used in the 
primary analysis to determine the longitudinal associations between the outcome 
variables and the predictor variables. The predictor variables
were scaled (mean-centered and standardized). Given the longitudinal 
design, an auto-regressive of order 1 working correlation matrix was 
specified in the GEE model. Covariates to adjust for were selected based on the
previous literature, from directed acyclic graph (DAG) [@Greenland1999a]
recommendations, and from quasi-likelihood information criteria. 
DAGs are used to identify the minimum adjustment necessary for a model by
using the causal pathways to algorithmically identify potential confounding and
colliding variables (see the [@Greenland1999a] ref for more detail about using
DAGs). The DAG structures to understand potential confounding, shown in 
`r cite_sf("dag_is")` and `r cite_sf("dag_bcf")`, were processed by the DAGitty
software [@Textor2011a;@Shrier2008a] to generate the recommended adjustments.
These DAG structures were developed based on hypothesized causal pathways
between each variable, which were then input into the DAGitty software. The
output from DAGitty was used to help inform the final model.

@@

There are a few possible explanations to these findings. Circulating 
TGFA derive from three sources: adipose lyposis, dietary fat, and de novo
lipogenesis (DNL). Dietary carbohydrates and fat can influence DNL activity
[@Aarsland1998a;@Harding2015a;@Luukkonen2018a]. Determining the specific source
of TGFA is extremely difficult to ascertain outside of highly controlled
experimental settings. Many previous studies that have examined DNL
have used markers of estimated DNL, such as the ratio between 18:2n-6 to 16:0 or
16:1n-7 to 16:0 [@Hodson2008a; @Lankinen2015a; @Kroger2011a]. However, there are
limitations to using these ratios as the fatty acids used in their calculation
can also be obtained from the diet in addition to being created through DNL
[@Hodson2008a]. An experimental feeding trial (n=24) was conducted to identify the fatty acids
that most accurately reflected DNL as potential biomarkers [@Lee2015a]. The
study found that palmitoleic acid (16:1n-7), directly measured DNL using
isotopes, and liver fat were all highly correlated with each other (r>0.50),
suggesting that 16:1n-7 may be a good biomarker for hepatic DNL. In another
small (n=14) feeding trial, meal type (high fat vs low fat) was tested to
determine its effect on DNL and TGFA composition [@Wilke2009a]. The authors
reported that 14:0, 16:0, 16:1, and 18:2 were higher in the low fat (high
carbohydrate) group. In another recent overfeeding trial, 1000 kcal of saturated
fat, unsaturated fat, or carbohydrates over 3 weeks was given to 38 overweight
individuals [@Luukkonen2018a] to test changes in liver fat and hepatic DNL. At the
end of the study, the carbohydrate group had higher DNL as well as an increase
in liver fat, though the saturated fat group had the highest increase in liver
fat.

@@

Regardless of the exact source of these fatty acids, our results in addition to
the available scientific evidence emphasizes the importance of the fatty acid
composition on metabolic health, as individual fatty acids can have specific
physiological functions. For instance, a higher concentration of
circulating 14 and 16 carbon fatty acids may expose tissues to greater
lipotoxicity, for instance from palmitic acid (16:0), which is well-known to
have harmful effects on tissues [@Riserus2008a; @Iggman2010a]. Our study extends
these findings by showing that TGFA with 14 to 16 carbons clustered together and
this pattern strongly associated with lower insulin sensitivity. While these fatty
acids also had a significant association with beta-cell function, the magnitude
of associations were more modest compared to those for insulin sensitivity.

@@

Given the close biological relationship between circulating NEFA and TG, NEFA 
may act as confounding factor and was thus adjusted for. In our published
analysis of the NEFA fraction [@Johnston2017b], we found that higher total NEFA,
but not the specific composition, associated with lower beta-cell function. This
is in contrast to the TGFA findings that the specific composition does
differentially associate with insulin sensitivity and beta-cell function,
adjusting for total NEFA. There was no difference in results in models that did
not include NEFA as a confounder (data not shown). This difference in results
between NEFA and TGFA suggests that TGFA may independently and strongly influence the 
pathophysiology of T2D, when compared to other lipid fraction compositions,
including the phospholipid and cholesteryl ester fractions [@Johnston2016a]. This
may be because TG are biologically destined for uptake by non-hepatic tissue as
they are found mainly in VLDL, at least during fasting. This is in contrast to
NEFA that are mostly uptaken by the liver and used in TG production [@Nielsen2012a].

@@

In conclusion, we found that a TGFA composition containing higher proportions of
14:0, 14:1n-7, 16:0, and 16:1n-7 associated strongly with lower insulin
sensitivity and (more moderately) with lower beta-cell function. We also found
that most other TGFA (e.g. 20:0, most omega 6 and 9 TGFA, and 20:5n-3)
associated positively with insulin sensitivity. Only a few TGFA associated
positively and consistently with beta-cell function (e.g. 18:1n-7, 20:2n-6). 
These results provide more insight into how individual TGFA contribute to the
pathogenesis of T2D while reinforcing the importance and value of clinically
measured total TG as a indicator of metabolic health.

>>>>

Generalized estimating equation (GEE) models [@Zeger1986a] were used in the 
primary analysis to determine the longitudinal associations between the outcome 
variables and the predictor variables. The predictor variables
were scaled (mean-centered and standardized). Given the longitudinal 
design, an auto-regressive of order 1 working correlation matrix was 
specified in the GEE model. Covariates to adjust for were selected based on the
previous literature, from directed acyclic graph (DAG) [@Greenland1999a]
recommendations, and from quasi-likelihood information criteria. 
DAGs are used to identify the minimum adjustment necessary for a model by
using the causal pathways to algorithmically identify potential confounding and
colliding variables (see [@Greenland1999a] for more detail about using
DAGs). The DAG structures to understand potential confounding, shown in 
`r cite_sf("dag_is")` and `r cite_sf("dag_bcf")`, were processed by the DAGitty
software [@Textor2011a;@Shrier2008a] to generate the recommended adjustments.
These DAG structures were developed based on hypothesized causal pathways
between each variable, which were then input into the DAGitty software. The
output from DAGitty was used to help inform the final model.

@@

There are a few possible explanations for these findings. Circulating 
TGFA derive from three sources: adipose lyposis, dietary fat, and de novo
lipogenesis (DNL). Dietary carbohydrates and fat can influence DNL activity
[@Aarsland1998a;@Harding2015a;@Luukkonen2018a]. Determining the specific source
of TGFA is extremely difficult to ascertain outside of highly controlled
experimental settings. Many previous studies that have examined DNL
have used markers of estimated DNL, such as the ratio between 18:2n-6 to 16:0 or
16:1n-7 to 16:0 [@Hodson2008a; @Lankinen2015a; @Kroger2011a]. However, there are
limitations to using these ratios as the fatty acids used in their calculation
can also be obtained from the diet in addition to being created through DNL
[@Hodson2008a]. An experimental feeding trial (n=24) was conducted to identify the fatty acids
that most accurately reflected DNL as potential biomarkers [@Lee2015a]. The
study found that palmitoleic acid (16:1n-7), directly measured DNL using
isotopes, and liver fat were all highly correlated with each other (r>0.50),
suggesting that 16:1n-7 may be a good biomarker for hepatic DNL. In another
small (n=14) feeding trial, meal type (high fat vs low fat) was tested to
determine its effect on DNL and TGFA composition [@Wilke2009a]. The authors
reported that 14:0, 16:0, 16:1, and 18:2 were higher in the low fat (high
carbohydrate) group. In another recent overfeeding trial, 1000 kcal of saturated
fat, unsaturated fat, or carbohydrates over 3 weeks was given to 38 overweight
individuals [@Luukkonen2018a] to test changes in liver fat and hepatic DNL. At the
end of the study, the carbohydrate group had higher DNL as well as an increase
in liver fat, though the saturated fat group had the highest increase in liver
fat.

@@

Regardless of the exact source of these fatty acids, our results, in addition to
the available scientific evidence, emphasize the importance of the fatty acid
composition on metabolic health, as individual fatty acids can have specific
physiological functions. For instance, a higher concentration of
circulating 14 and 16 carbon fatty acids may expose tissues to greater
lipotoxicity, for instance from palmitic acid (16:0), which is well-known to
have harmful effects on tissues [@Riserus2008a; @Iggman2010a]. Our study extends
these findings by showing that TGFA with 14 to 16 carbons clustered together and
this pattern strongly associated with lower insulin sensitivity. While these fatty
acids also had a significant association with beta-cell function, the magnitude
of associations were more modest compared to those for insulin sensitivity.

@@

Given the close biological relationship between circulating NEFA and TG, NEFA 
may act as confounding factor and was thus adjusted for. In our published
analysis of the NEFA fraction [@Johnston2017b], we found that higher total NEFA,
but not the specific composition, associated with lower beta-cell function. This
is in contrast to the TGFA findings that the specific composition does
differentially associate with insulin sensitivity and beta-cell function,
adjusting for total NEFA. There was no difference in results in models that did
not include NEFA as a confounder (data not shown). This difference in results
between NEFA and TGFA suggests that TGFA may independently and strongly influence the 
pathophysiology of T2D, when compared to other lipid fraction compositions,
including the phospholipid and cholesteryl ester fractions [@Johnston2016a]. This
may be because TG are biologically destined for uptake by non-hepatic tissue as
they are found mainly in VLDL, at least during fasting. This is in contrast to
NEFA that are mostly taken up by the liver and used in TG production [@Nielsen2012a].

@@

In conclusion, we found that a TGFA composition containing higher proportions of
14:0, 14:1n-7, 16:0, and 16:1n-7 associated strongly with lower insulin
sensitivity and (more moderately) with lower beta-cell function. We also found
that most other TGFA (e.g. 20:0, most omega 6 and 9 TGFA, and 20:5n-3)
associated positively with insulin sensitivity. Only a few TGFA associated
positively and consistently with beta-cell function (e.g. 18:1n-7, 20:2n-6). 
These results provide more insight into how individual TGFA contribute to the
pathogenesis of T2D while reinforcing the importance and value of clinically
measured total TG as an indicator of metabolic health.
